Generalized Plackett-Luce Likelihoods

### 2022-08-16 inst/jss_hyper3.Rnw
slight modifications including citing hankin 2010

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
are integers which may be positive or negative.  It becomes possible
to analyse a wider range of likelihood functions than standard
Plackett-Luce~\citep{turner2020}.  For example, results involving
incomplete order statistics or teams are tractable. Further, the
introduction of reified entities allows one to consider
draws~\citep{hankin2009}, noncompetitive tactics such as
collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its parts.

@@

Here I consider some order statistics with nontrivial maximum
likelihood Bradley-Terry strengths.  The simplest nontrivial case
would be three competitors with strengths $a,a,b$ and finishing order
$a\succ b\succ a$.  The likelihood function would be

@@

In this case we know that $a+b=1$ so this is equal to
$\mathcal{L}=\mathcal{L}(a)=\frac{a(1-a)}{1+a}$.

@@

This will be zero at $\sqrt{2}-1$; we also note that
$d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
$0\leq a\leq 1$: the root is a maximum.

@@

which has a minimum of about $13.9$ at at about $a=0.575$.

@@

The ideas presented above can easily be extended to arbitrarily large
numbers of competitors, although the analytical expressions tend to be
intractable and numerical methods must be used.  All results and
datasets presented here are maintained under version control and
available at \url{https:github.com/RobinHankin/hyper2}.  Given an
order statistic of the type considered above, the
Mann-Whitney-Wilcoxon test~\citep{mann1947,wilcoxon1945} assesses a
null of identity of underlying distributions.  Consider the
chorioamnion dataset~\citep{hollander2013}, used in
\code{wilcox.test.Rd}:

@@

Draws are not considered in the present work but in principle may be
accommodated, either using likelihoods comprising sums of
Plackett-Luce probabilities~\citep{hankin2017}; or the introduction of
a reified draw entity~\citep{hankin2020}.

>>>>

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
are integers which may be positive or negative.  It becomes possible
to analyse a wider range of likelihood functions than standard
Plackett-Luce~\citep{turner2020}.  For example, results involving
incomplete order statistics or teams are tractable. Further, the
introduction of reified entities allows one to consider
draws~\citep{hankin2009}, noncompetitive tactics such as
collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its
parts~\citep{hankin2010}.

@@

Here I consider some order statistics with nontrivial maximum
likelihood Bradley-Terry strengths.  The simplest nontrivial case
would be three competitors with strengths $a,a,b$ and finishing order
$a\succ b\succ a$.  The Plackett-Luce likelihood function would be

@@

and in this case we know that $a+b=1$ so this is equal to
$\mathcal{L}=\mathcal{L}(a)=\frac{a(1-a)}{1+a}$.  The score would be
given by

@@

and this will be zero at $\sqrt{2}-1$; we also note that
$d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
$0\leq a\leq 1$: the root is a maximum.

@@

which has a minimum of about $13.9$ at at about $a=0.575$.  This is
intermediate between the two-runner case where the Fisher information
exceeds 4 (minimum at $p_a=\frac{1}{2}$), and the case of three
distinct runners, where the Fisher information exceeds
$\frac{1323}{16}\simeq 82.7$ (minimum at $p_a=p_b=p_c=\frac{1}{3}$); a
brief discussion is given at \code{inst/fisher\_inf\_PL3.Rmd}.

@@

The ideas presented above can easily be extended to arbitrarily large
numbers of competitors, although the analytical expressions tend to be
intractable and numerical methods must be used.  All results and
datasets presented here are maintained under version control and
available at \url{https:github.com/RobinHankin/hyper2}.  Given an
order statistic of the type considered above, the
Mann-Whitney-Wilcoxon test~\citep{mann1947,wilcoxon1945} assesses a
null of identity of underlying distributions~\citep{ahmad1996}.
Consider the chorioamnion dataset~\citep{hollander2013}, used in
\code{wilcox.test.Rd}:

@@

Draws are not considered in the present work but in principle may be
accommodated, either using likelihoods comprising sums of
Plackett-Luce probabilities~\citep{hankin2017}; or the introduction of
a reified draw entity~\citep{hankin2010}.
